A Bioconductor workflow for processing and analysing spatial proteomics data

### 2018-04-17 bioc-workflow-resubmit.Rnw
update unsup ml

<<writehl, eval=FALSE>>=
fData(hl)$pd <- as.character(fData(hl)$pd)
write.exprs(hl, fDataCols = "pd", file = "pd-results.csv", sep = ",")
@

@@

In \Biocpkg{pRoloc} there is functionality for unsupervsied machine
learning methods. In unsupervised learning, the training data consists
of a set of input vectors e.g. protein profiles, ignoring the
information about the class label e.g. localisation, other than for
annotation purposes. The main goal in unsupervised learning is to
uncover groups of similar features within the data, termed
clustering. Ordination methods such as principal components analysis
(PCA) also fall into the category of unsupervised learning methods,
where the data can be projected from a high-dimensional space down to
two or three dimensions for the purpose of visualisation.

As described and demonstrated already above, PCA is a valuable and
powerful method for data visualisation and quality control. Another
application uses hierarchical clustering to summarise the relation
between marker proteins using he \texttt{mrkHClust} function, where
the euclidean distance between average class-specific profiles is used
to produce a dendrogramme describing a simple relationship between the
sub-cellular classes (Figure~\ref{fig:mrkHclust}). The
\texttt{mrkHClust} uses the same defaults as all other function, using
the \texttt{markers} feature variable to define marker proteins. In
the code chunk, we adapt the figure margins to fully display the class
names.

>>>>

<<writehl, eval=FALSE>>=
fData(hl)$pd <- as.character(fData(hl)$pd)
write.exprs(hl, fcol = "pd", file = "pd-results.csv", sep = ",")
@

@@

In \Biocpkg{pRoloc} there is functionality for unsupervsied machine
learning methods. In unsupervised learning, the training data consists
of a set of input vectors e.g. protein profiles, ignoring the
information about the class label e.g. localisation, other than for
annotation purposes. The main goal in unsupervised learning is to
uncover groups of similar features within the data, termed
clustering. Ordination methods such as principal components analysis
(PCA) also fall into the category of unsupervised learning methods,
where the data can be projected from a high-dimensional space down to
two or three dimensions.

As described and demonstrated already above, PCA is a valuable and
powerful method for data visualisation and quality control. Another
application uses hierarchical clustering to summarise the relation
between marker proteins using the \texttt{mrkHClust} function, where
the euclidean distance between average class-specific profiles is used
to produce a dendrogramme describing a simple relationship between the
sub-cellular classes (Figure~\ref{fig:mrkHclust}). The
\texttt{mrkHClust} uses the same defaults as all other function, using
the \texttt{markers} feature variable to define marker proteins. In
the code chunk, we adapt the figure margins to fully display the class
names.

### 2018-04-17 bioc-workflow-resubmit.tex
update unsup ml

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{combined} \hlkwb{<-} \hlkwd{combine}\hlstd{(hyperLOPIT2015ms3r1, hyperLOPIT2015ms3r2)}
\hlstd{combined}
\end{alltt}
\begin{verbatim}
## MSnSet (storageMode: lockedEnvironment)
## assayData: 6725 features, 20 samples 
##   element names: exprs 
## protocolData: none
## phenoData
##   sampleNames: X126.1 X127N.1 ... X131.2 (20 total)
##   varLabels: Replicate TMT.Reagent ... Iodixonal.Density (5 total)
##   varMetadata: labelDescription
## featureData
##   featureNames: Q9JHU4 Q9QXS1-3 ... Q9Z2Y3-3 (6725 total)
##   fvarLabels: EntryName ProteinDescription ... ProteinCoverage2 (8
##     total)
##   fvarMetadata: labelDescription
## experimentData: use 'experimentData(object)'
## Annotation:  
## - - - Processing information - - -
## Combined [6725,20] and [6268,10] MSnSets Tue Apr 17 15:22:42 2018 
##  MSnbase version: 1.21.7
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{combined} \hlkwb{<-} \hlkwd{filterNA}\hlstd{(combined)}
\hlstd{combined}
\end{alltt}
\begin{verbatim}
## MSnSet (storageMode: lockedEnvironment)
## assayData: 5032 features, 20 samples 
##   element names: exprs 
## protocolData: none
## phenoData
##   sampleNames: X126.1 X127N.1 ... X131.2 (20 total)
##   varLabels: Replicate TMT.Reagent ... Iodixonal.Density (5 total)
##   varMetadata: labelDescription
## featureData
##   featureNames: Q9JHU4 Q9QXS1-3 ... Q9Z2R6 (5032 total)
##   fvarLabels: EntryName ProteinDescription ... ProteinCoverage2 (8
##     total)
##   fvarMetadata: labelDescription
## experimentData: use 'experimentData(object)'
## Annotation:  
## - - - Processing information - - -
## Combined [6725,20] and [6268,10] MSnSets Tue Apr 17 15:22:42 2018 
## Subset [6725,20][5032,20] Tue Apr 17 15:22:43 2018 
## Removed features with more than 0 NAs: Tue Apr 17 15:22:43 2018 
## Dropped featureData's levels Tue Apr 17 15:22:43 2018 
##  MSnbase version: 1.21.7
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{figure}[!ht]
  \centering
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{prots} \hlkwb{<-} \hlkwd{c}\hlstd{(}\hlstr{"Q8CG48"}\hlstd{,} \hlstr{"Q8CG47"}\hlstd{,} \hlstr{"Q8K2Z4"}\hlstd{,} \hlstr{"Q8C156"}\hlstd{)}
\hlstd{foi13s} \hlkwb{<-} \hlkwd{FeaturesOfInterest}\hlstd{(}\hlkwc{description} \hlstd{=} \hlstr{"13S consensin proteins"}\hlstd{,}
                             \hlkwc{fnames} \hlstd{= prots,}
                             \hlkwc{object} \hlstd{= hl)}
\hlstd{foi13s}
\end{alltt}
\begin{verbatim}
## Traceable object of class "FeaturesOfInterest"
##  Created on Tue Apr 17 15:22:45 2018 
##  Description:
##   13S consensin proteins
##  4 features of interest:
##    Q8CG48, Q8CG47, Q8K2Z4, Q8C156
\end{verbatim}
\begin{alltt}
\hlkwd{plot2D}\hlstd{(hl)}
\hlkwd{addLegend}\hlstd{(hl,} \hlkwc{cex} \hlstd{=} \hlnum{.6}\hlstd{)}
\hlkwd{highlightOnPlot}\hlstd{(hl, foi13s)}
\end{alltt}
\end{kframe}
\includegraphics[width=.8\textwidth]{figure/foi-1}

@@

% latex table generated in R 3.4.4 by xtable 1.8-2 package
% Tue Apr 17 15:22:45 2018
\begin{table}[ht]
\centering
\begin{tabular}{rrllr}
  \hline
 & Replicate & Tag & Gradient.Fraction & Iodixonal.Density \\ 
  \hline
X127C &   1 & 127C & 8 to 9 (pooled) & 11.00 \\ 
  X127C.1 &   2 & 127C & 7 to 9 (pooled) & 10.00 \\ 
   \hline
\end{tabular}
\caption{Differences in gradient fraction pooling.} 
\label{tab:pdtab}
\end{table}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{hl}
\end{alltt}
\begin{verbatim}
## MSnSet (storageMode: lockedEnvironment)
## assayData: 5032 features, 20 samples 
##   element names: exprs 
## protocolData: none
## phenoData
##   sampleNames: X126 X127N ... X131.1 (20 total)
##   varLabels: Replicate Tag Gradient.Fraction Iodixonal.Density
##   varMetadata: labelDescription
## featureData
##   featureNames: Q9JHU4 Q9QXS1-3 ... Q9Z2R6 (5032 total)
##   fvarLabels: uniprot.accession uniprot.id ... pd (27 total)
##   fvarMetadata: labelDescription
## experimentData: use 'experimentData(object)'
## Annotation:  
## - - - Processing information - - -
## Added markers from  'mrk' marker vector. Tue Apr 17 15:22:45 2018 
## Added markers from  'pdres' marker vector. Tue Apr 17 15:22:46 2018 
##  MSnbase version: 2.4.2
\end{verbatim}
\begin{alltt}
\hlkwd{getMarkers}\hlstd{(hl,} \hlkwc{fcol} \hlstd{=} \hlstr{"pd"}\hlstd{)}
\end{alltt}
\begin{verbatim}
## organelleMarkers
##                          40S Ribosome 
##                                   106 
##                          60S Ribosome 
##                                    95 
## Endoplasmic reticulum/Golgi apparatus 
##                                   393 
##                         Mitochondrion 
##                                   525 
##                           Phenotype 1 
##                                   300 
##                           Phenotype 2 
##                                   253 
##                           Phenotype 3 
##                                   203 
##                           Phenotype 4 
##                                    74 
##                           Phenotype 5 
##                                    91 
##                       Plasma membrane 
##                                   421 
##                            Proteasome 
##                                    92 
##                               unknown 
##                                  2479
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlkwd{fData}\hlstd{(hl)}\hlopt{$}\hlstd{pd} \hlkwb{<-} \hlkwd{as.character}\hlstd{(}\hlkwd{fData}\hlstd{(hl)}\hlopt{$}\hlstd{pd)}
\hlkwd{write.exprs}\hlstd{(hl,} \hlkwc{fDataCols} \hlstd{=} \hlstr{"pd"}\hlstd{,} \hlkwc{file} \hlstd{=} \hlstr{"pd-results.csv"}\hlstd{,} \hlkwc{sep} \hlstd{=} \hlstr{","}\hlstd{)}
\end{alltt}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{(ts} \hlkwb{<-} \hlkwd{orgQuants}\hlstd{(hl,} \hlkwc{fcol} \hlstd{=} \hlstr{"svm"}\hlstd{,} \hlkwc{scol} \hlstd{=} \hlstr{"svm.scores"}\hlstd{,} \hlkwc{mcol} \hlstd{=} \hlstr{"markers"}\hlstd{,} \hlkwc{t} \hlstd{=} \hlnum{.5}\hlstd{))}
\end{alltt}
\begin{verbatim}
##            40S Ribosome            60S Ribosome      Actin cytoskeleton 
##               0.4323939               0.3044486               0.3845760 
##                 Cytosol   Endoplasmic reticulum                Endosome 
##               0.6850826               0.6012822               0.4245089 
##    Extracellular matrix                Lysosome           Mitochondrion 
##               0.4256415               0.5944635               0.9493494 
##     Nucleus - Chromatin Nucleus - Non-chromatin              Peroxisome 
##               0.7949812               0.7104192               0.3114933 
##         Plasma membrane              Proteasome 
##               0.7229559               0.4157514
##            40S Ribosome            60S Ribosome      Actin cytoskeleton 
##               0.4323939               0.3044486               0.3845760 
##                 Cytosol   Endoplasmic reticulum                Endosome 
##               0.6850826               0.6012822               0.4245089 
##    Extracellular matrix                Lysosome           Mitochondrion 
##               0.4256415               0.5944635               0.9493494 
##     Nucleus - Chromatin Nucleus - Non-chromatin              Peroxisome 
##               0.7949812               0.7104192               0.3114933 
##         Plasma membrane              Proteasome 
##               0.7229559               0.4157514
\end{verbatim}
\begin{alltt}
\hlstd{hl} \hlkwb{<-} \hlkwd{getPredictions}\hlstd{(hl,} \hlkwc{fcol} \hlstd{=} \hlstr{"svm"}\hlstd{,} \hlkwc{scol} \hlstd{=} \hlstr{"svm.scores"}\hlstd{,} \hlkwc{mcol} \hlstd{=} \hlstr{"markers"}\hlstd{,} \hlkwc{t} \hlstd{= ts)}
\end{alltt}
\begin{verbatim}
## ans
##            40S Ribosome            60S Ribosome      Actin cytoskeleton 
##                      85                     170                      88 
##                 Cytosol   Endoplasmic reticulum                Endosome 
##                     296                     475                     101 
##    Extracellular matrix                Lysosome           Mitochondrion 
##                      27                     124                     524 
##     Nucleus - Chromatin Nucleus - Non-chromatin              Peroxisome 
##                     230                     343                      38 
##         Plasma membrane              Proteasome                 unknown 
##                     316                     158                    2057
\end{verbatim}
\end{kframe}
\end{knitrout}

>>>>

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{combined} \hlkwb{<-} \hlkwd{combine}\hlstd{(hyperLOPIT2015ms3r1, hyperLOPIT2015ms3r2)}
\hlstd{combined}
\end{alltt}
\begin{verbatim}
## MSnSet (storageMode: lockedEnvironment)
## assayData: 6725 features, 20 samples 
##   element names: exprs 
## protocolData: none
## phenoData
##   sampleNames: X126.1 X127N.1 ... X131.2 (20 total)
##   varLabels: Replicate TMT.Reagent ... Iodixonal.Density (5 total)
##   varMetadata: labelDescription
## featureData
##   featureNames: Q9JHU4 Q9QXS1-3 ... Q9Z2Y3-3 (6725 total)
##   fvarLabels: EntryName ProteinDescription ... ProteinCoverage2 (8
##     total)
##   fvarMetadata: labelDescription
## experimentData: use 'experimentData(object)'
## Annotation:  
## - - - Processing information - - -
## Combined [6725,20] and [6268,10] MSnSets Tue Apr 17 15:40:33 2018 
##  MSnbase version: 1.21.7
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{combined} \hlkwb{<-} \hlkwd{filterNA}\hlstd{(combined)}
\hlstd{combined}
\end{alltt}
\begin{verbatim}
## MSnSet (storageMode: lockedEnvironment)
## assayData: 5032 features, 20 samples 
##   element names: exprs 
## protocolData: none
## phenoData
##   sampleNames: X126.1 X127N.1 ... X131.2 (20 total)
##   varLabels: Replicate TMT.Reagent ... Iodixonal.Density (5 total)
##   varMetadata: labelDescription
## featureData
##   featureNames: Q9JHU4 Q9QXS1-3 ... Q9Z2R6 (5032 total)
##   fvarLabels: EntryName ProteinDescription ... ProteinCoverage2 (8
##     total)
##   fvarMetadata: labelDescription
## experimentData: use 'experimentData(object)'
## Annotation:  
## - - - Processing information - - -
## Combined [6725,20] and [6268,10] MSnSets Tue Apr 17 15:40:33 2018 
## Subset [6725,20][5032,20] Tue Apr 17 15:40:34 2018 
## Removed features with more than 0 NAs: Tue Apr 17 15:40:34 2018 
## Dropped featureData's levels Tue Apr 17 15:40:34 2018 
##  MSnbase version: 1.21.7
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{figure}[!ht]
  \centering
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{prots} \hlkwb{<-} \hlkwd{c}\hlstd{(}\hlstr{"Q8CG48"}\hlstd{,} \hlstr{"Q8CG47"}\hlstd{,} \hlstr{"Q8K2Z4"}\hlstd{,} \hlstr{"Q8C156"}\hlstd{)}
\hlstd{foi13s} \hlkwb{<-} \hlkwd{FeaturesOfInterest}\hlstd{(}\hlkwc{description} \hlstd{=} \hlstr{"13S consensin proteins"}\hlstd{,}
                             \hlkwc{fnames} \hlstd{= prots,}
                             \hlkwc{object} \hlstd{= hl)}
\hlstd{foi13s}
\end{alltt}
\begin{verbatim}
## Traceable object of class "FeaturesOfInterest"
##  Created on Tue Apr 17 15:40:38 2018 
##  Description:
##   13S consensin proteins
##  4 features of interest:
##    Q8CG48, Q8CG47, Q8K2Z4, Q8C156
\end{verbatim}
\begin{alltt}
\hlkwd{plot2D}\hlstd{(hl)}
\hlkwd{addLegend}\hlstd{(hl,} \hlkwc{cex} \hlstd{=} \hlnum{.6}\hlstd{)}
\hlkwd{highlightOnPlot}\hlstd{(hl, foi13s)}
\end{alltt}
\end{kframe}
\includegraphics[width=.8\textwidth]{figure/foi-1}

@@

% latex table generated in R 3.4.4 by xtable 1.8-2 package
% Tue Apr 17 15:40:38 2018
\begin{table}[ht]
\centering
\begin{tabular}{rrllr}
  \hline
 & Replicate & Tag & Gradient.Fraction & Iodixonal.Density \\ 
  \hline
X127C &   1 & 127C & 8 to 9 (pooled) & 11.00 \\ 
  X127C.1 &   2 & 127C & 7 to 9 (pooled) & 10.00 \\ 
   \hline
\end{tabular}
\caption{Differences in gradient fraction pooling.} 
\label{tab:pdtab}
\end{table}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{hl}
\end{alltt}
\begin{verbatim}
## MSnSet (storageMode: lockedEnvironment)
## assayData: 5032 features, 20 samples 
##   element names: exprs 
## protocolData: none
## phenoData
##   sampleNames: X126 X127N ... X131.1 (20 total)
##   varLabels: Replicate Tag Gradient.Fraction Iodixonal.Density
##   varMetadata: labelDescription
## featureData
##   featureNames: Q9JHU4 Q9QXS1-3 ... Q9Z2R6 (5032 total)
##   fvarLabels: uniprot.accession uniprot.id ... pd (27 total)
##   fvarMetadata: labelDescription
## experimentData: use 'experimentData(object)'
## Annotation:  
## - - - Processing information - - -
## Added markers from  'mrk' marker vector. Tue Apr 17 15:40:37 2018 
## Added markers from  'pdres' marker vector. Tue Apr 17 15:40:39 2018 
##  MSnbase version: 2.4.2
\end{verbatim}
\begin{alltt}
\hlkwd{getMarkers}\hlstd{(hl,} \hlkwc{fcol} \hlstd{=} \hlstr{"pd"}\hlstd{)}
\end{alltt}
\begin{verbatim}
## organelleMarkers
##                          40S Ribosome 
##                                   106 
##                          60S Ribosome 
##                                    95 
## Endoplasmic reticulum/Golgi apparatus 
##                                   393 
##                         Mitochondrion 
##                                   525 
##                           Phenotype 1 
##                                   300 
##                           Phenotype 2 
##                                   253 
##                           Phenotype 3 
##                                   203 
##                           Phenotype 4 
##                                    74 
##                           Phenotype 5 
##                                    91 
##                       Plasma membrane 
##                                   421 
##                            Proteasome 
##                                    92 
##                               unknown 
##                                  2479
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlkwd{fData}\hlstd{(hl)}\hlopt{$}\hlstd{pd} \hlkwb{<-} \hlkwd{as.character}\hlstd{(}\hlkwd{fData}\hlstd{(hl)}\hlopt{$}\hlstd{pd)}
\hlkwd{write.exprs}\hlstd{(hl,} \hlkwc{fcol} \hlstd{=} \hlstr{"pd"}\hlstd{,} \hlkwc{file} \hlstd{=} \hlstr{"pd-results.csv"}\hlstd{,} \hlkwc{sep} \hlstd{=} \hlstr{","}\hlstd{)}
\end{alltt}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{(ts} \hlkwb{<-} \hlkwd{orgQuants}\hlstd{(hl,} \hlkwc{fcol} \hlstd{=} \hlstr{"svm"}\hlstd{,} \hlkwc{scol} \hlstd{=} \hlstr{"svm.scores"}\hlstd{,} \hlkwc{mcol} \hlstd{=} \hlstr{"markers"}\hlstd{,} \hlkwc{t} \hlstd{=} \hlnum{.5}\hlstd{))}
\end{alltt}
\begin{verbatim}
##            40S Ribosome            60S Ribosome      Actin cytoskeleton 
##               0.4382130               0.3040042               0.3797701 
##                 Cytosol   Endoplasmic reticulum                Endosome 
##               0.6897383               0.6007307               0.4284784 
##    Extracellular matrix                Lysosome           Mitochondrion 
##               0.4200181               0.5879109               0.9502066 
##     Nucleus - Chromatin Nucleus - Non-chromatin              Peroxisome 
##               0.7951514               0.7100151               0.3166095 
##         Plasma membrane              Proteasome 
##               0.7050611               0.4169158
##            40S Ribosome            60S Ribosome      Actin cytoskeleton 
##               0.4382130               0.3040042               0.3797701 
##                 Cytosol   Endoplasmic reticulum                Endosome 
##               0.6897383               0.6007307               0.4284784 
##    Extracellular matrix                Lysosome           Mitochondrion 
##               0.4200181               0.5879109               0.9502066 
##     Nucleus - Chromatin Nucleus - Non-chromatin              Peroxisome 
##               0.7951514               0.7100151               0.3166095 
##         Plasma membrane              Proteasome 
##               0.7050611               0.4169158
\end{verbatim}
\begin{alltt}
\hlstd{hl} \hlkwb{<-} \hlkwd{getPredictions}\hlstd{(hl,} \hlkwc{fcol} \hlstd{=} \hlstr{"svm"}\hlstd{,} \hlkwc{scol} \hlstd{=} \hlstr{"svm.scores"}\hlstd{,} \hlkwc{mcol} \hlstd{=} \hlstr{"markers"}\hlstd{,} \hlkwc{t} \hlstd{= ts)}
\end{alltt}
\begin{verbatim}
## ans
##            40S Ribosome            60S Ribosome      Actin cytoskeleton 
##                      86                     169                      85 
##                 Cytosol   Endoplasmic reticulum                Endosome 
##                     296                     477                      94 
##    Extracellular matrix                Lysosome           Mitochondrion 
##                      27                     123                     522 
##     Nucleus - Chromatin Nucleus - Non-chromatin              Peroxisome 
##                     230                     344                      39 
##         Plasma membrane              Proteasome                 unknown 
##                     325                     157                    2058
\end{verbatim}
\end{kframe}
\end{knitrout}
